---
title: "1. Pathway Visualization (WikiPathways)"
author: "Jarno Koetsier"
package: PinPath
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
    toc_float: true
    includes:
      in_header: Pathway-Visualization-schema.html
# pdf_document:
#   toc: true
vignette: >
  %\VignetteIndexEntry{Pathway Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(tidy = FALSE,
                      cache = FALSE,
                      dev = "png",
                      message = FALSE, error = FALSE, warning = TRUE)
```

This vignette will cover visualizing differential expression analysis results onto *WikiPathways* pathway diagrams.

# Installation
First, make sure you have *PinPath* installed.
```{r, message = FALSE, warning = FALSE}
library(PinPath)
library(rWikiPathways)
library(org.Hs.eg.db)
```


# Dataset
The example dataset we will use compares the expression of transcripts in lung cancer biopsies versus normal tissue. Differential expression analysis has already been performed, generating log2FCs and p-values for each gene. 
```{r, message = FALSE, warning = FALSE}
lung_expr <- read.csv(system.file("extdata","data-lung-cancer.csv", package="PinPath"), 
                      stringsAsFactors = FALSE)
```

In the pathway, we want to show which gene is significantly differentially expressed. For this, we will use an adjusted p-value cutoff of 0.05.
```{r, results = 'asis'}
lung_expr$Significant <- ifelse(lung_expr$adj.P.Value < 0.05, "Yes", "No")
```

# Set colors
After we have loaded and prepared the differential expression analysis statistics, we need to define how to color the statistics in the pathway diagram. In this vignette, we will plot the log2FC and significance of each gene on the pathway diagram. We can start by loading the default color palette.
```{r}
colorList <- PinPath::defaultColorList(lung_expr[,c("log2FC", "Significant")])
```

In the next step, we can adjust the default color palette to the desired color values. For instance, we want to display a green color when a gene is differentially expressed  and a white color when it is not.
```{r}
colorList[["Significant"]]$Color <- c("Yes" = "green",
                                      "No" = "white")
```

Furthermore, we can set the minimum and maximum value  of the log2FC color gradient to -1.5 and 1.5, respectively.
```{r}
colorList[["log2FC"]]$ColorVal <- c("MinVal" = -1.5,
                                    "MidVal" = 0,
                                    "MaxVal" = 1.5)
```

# Plot pathway
We can now plot the differential expression statistics on the *WP5087: Pleural mesothelioma* pathway.
If not specified otherwise, the pathway and legend image will be saved in your working directory. 
The pathway image will be opened by default.
```{r, message = FALSE, warning = FALSE}
# WP5087: Pleural mesothelioma
pathway_id <- "WP5087"
infile <- rWikiPathways::getPathway(pathway_id)

# Draw pathway
pathVis <- PinPath::drawGPML(
  infile = infile,
  annGenes = "org.Hs.eg.db",
  inputDB = "ENSEMBL",
  geneIDs = lung_expr$GeneID,
  colorVar = lung_expr[,c("log2FC", "Significant")],
  colorList = colorList,
  nodeTable = TRUE,
  legend = TRUE,
)
```

Pathway image:
<img src="Pleural_mesothelioma_WP5087_r140461_Homo_sapiens.svg" width="1000">

Legend image:
<img src="legend_Pleural_mesothelioma_WP5087_r140461_Homo_sapiens.svg" width="500">

# Session info

```{r sessionInfo}
sessionInfo()
```

